---
title: "Bicycle Rentals exploration by Sreejith Gopinath"
output:
  html_document:
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
    theme: united
    highlight: tango
    fig_caption: true
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```
# Univariate plots section

```{r, echo=FALSE}
setwd('/Users/sreeg/Exploratory_data_analysis')
df <- read.csv('hour.csv')
df_daily <- read.csv('day.csv')
df_master <- read.csv('hour.csv')

hr_casreg <- subset(df, select=c("hr","casual","registered"))
df <- subset(df, select = -c(casual, registered))
```

<span style="color:black">These are the dimensions of the bicycle_rentals dataset</span>

```{r, echo=FALSE}
dim(df)
```

<span style="color:black">The features in this dataset are:</span>
```{r, echo=FALSE}
colnames(df)

#Change values of certain columns from mere numbers to meaningful words
df$season[df$season == 1] <- "sp"
df$season[df$season == 2] <- "su"
df$season[df$season == 3] <- "fa"
df$season[df$season == 4] <- "wi"

df$weathersit[df$weathersit == 1] <- "cl"
df$weathersit[df$weathersit == 2] <- "mi"
df$weathersit[df$weathersit == 3] <- "light"
df$weathersit[df$weathersit == 4] <- "heavy"

#Print out first few records
head(df)

#Print out unique values of a few columns
unique(df$weathersit)
unique(df$season)

str(df)

summary(df)
```

<span style="color:black">
We see that the minimum ridership values on any given day is 1 and the maximum on any given day is just shy of 1000.</span>

```{r, echo=FALSE}
lowest_ridership <- subset(df_daily[order(df_daily$cnt, decreasing=F),])
lowest_ridership <- lowest_ridership[,c("dteday","cnt")]
head(lowest_ridership)
```

<span style="color:black">We see from the preceding dataframe dump that the lowest ever bicycle rentals in our dataset happened on 29th October 2012. A simple web search shows us that this was the day Hurricane Sandy made landfall in the Washington DC area - a majority of schools, colleges and offices had shut down! The next low occurred on 27th January 2011. On this day, it was a blizzard that had shut down Washington DC! Thus, analyzing transportation statistics over a period of time and co-relating incidents that correspond to variations to the trend, may help us build predictive models.</span>

```{r, echo=FALSE}
library(ggplot2)
library(reshape2)
library(GGally)
library(scales)
library(memisc)
library(lattice)
library(MASS)
library(plyr)
ggplot(aes(cnt), data=df) + geom_histogram(binwidth=10, position="identity", fill="#006633") + scale_y_log10() + ggtitle("Total bicycle rentals")
```

<span style="color:black">The bicycle rental frequencies seem to progressively decline for higher counts. I wonder how the frequencies will vary by season and weather conditions.</span>

```{r, echo=FALSE}
ggplot(aes(x=weathersit, y=cnt), data=df) + stat_summary(fun.y=mean, geom="bar", fill="#006633") + ggtitle("Total bicycle rentals by prevailing weather situation")
```

<span style="color:black">As is intuitive, we see that bicycle rentals are highest when the weather is clear, next highest when the weather is just misty, a few brave souls renting bicycles when there is light precipitation and least when there is heavy precipitation.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=cnt), data=df) + stat_summary(fun.y=mean, geom="bar", fill="#006633") + ggtitle("Median hourly total bicycle rentals on holidays")
```

<span style="color:black">I cannot fathom the reason for this, but it looks like bicycle rentals are highest in the fall, next highest in summer, third in winter and lowest in spring.</span>

<span style="color:black">Next, I subset-ed the data into two groups - one subset for holidays and one subset for working days. Here are the summaries of the bicycle rental numbers for the respective subsets:</span>

```{r, echo=FALSE}
df_working <- subset(df, holiday == 0)
df_holiday <- subset(df, holiday == 1)
summary(df_working$cnt)
summary(df_holiday$cnt)
```

<span style="color:black">From the preceding stats, we see that, all else set aside, bicycle rental numbers are higher on working days than on holidays - indicating that bicycle rentals may not be for recreational use.</span>

<span style="color:black">Next, let's explore how bicycle rentals stack up by day of the week.</span>

```{r, echo=FALSE}
ggplot(aes(x=weekday, y=cnt), data=df) + stat_summary(fun.y=median, geom="line", color="red") + ggtitle("Median daily total bicycle rentals")
```

<span style="color:black">From the preceding plot, we see that the busiest day for bicycle rentals is Friday, and the least busy day is Sunday. Rental numbers climb up through the work week and then fall off on the weekend.</span>

<span style="color:black">Let's now look at how bicycle rental numbers stack up by hour of the day.</span>

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt), data=df) + scale_x_continuous(breaks=-1:24) + stat_summary(fun.y=median, geom="line", color="red") + ggtitle("Median hourly total bicycle rentals")
```

<span style="color:black">We see that bicycle rentals peak at 8:00am and 5:00pm and drop off on either side, suggesting that the peaks may be driven up by people using bicycles to get to and from work.</span>

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt), data=df_working) + scale_x_continuous(breaks=-1:24) + stat_summary(fun.y=median, geom="line", color="red") + ggtitle("Median hourly total bicycle rentals on working days")
ggplot(aes(x=hr, y=cnt), data=df_holiday) + scale_x_continuous(breaks=-1:24) + stat_summary(fun.y=median, geom="line", color="red") + ggtitle("Median hourly total bicycle rentals on holidays")
```

<span style="color:black">The preceding two plots are of bicycle rentals by the hour on working days and by the hour on holidays. While the plot for holidays shows higher numbers during the middle portion of the day, the plot for hourly rentals on working days is shockingly similar to the plot for the overall dataset - further reinforcing our belief that most bicycle rentals are made by people getting to-and-from work/school.</span>

<span style="color:black">Let us now see how the numbers of casual and registered renters vary by the day of the week.</span>

```{r, echo=FALSE}
day_casreg <- subset(df_daily, select=c("weekday","casual","registered"))
day_casreg <- melt(day_casreg, id.vars="weekday")
ggplot(day_casreg, aes(weekday,value, col=variable)) + stat_summary(fun.y=median, geom="line", color="red") + ggtitle("Median daily bicycle rentals")
```

<span style="color:black">Let us now see how the numbers of casual and registered renters vary by hour of the day.</span>

```{r, echo=FALSE}
hr_casreg <- melt(hr_casreg, id.vars="hr")
ggplot(hr_casreg, aes(hr,value, col=variable)) + stat_summary(fun.y=median, geom="line", color="red") + ggtitle("Median hourly bicycle rentals")
```

<span style="color:black">From the preceding graphs, we see that casual ridership is higher on weekends than on weekdays, while the trend is reversed for registered ridership. We also see that the hourly ridership plot for registered riders is similar to the overall hourly ridership plot - which means that rental registration is a pretty good predictor of bicycle rentals.</span>

# Univariate analysis

<span style="color:black"><b><i>What is the structure of your dataset?</i></b></span><br>
<span style="color:red">I am working with the bicycle rentals dataset, which contains bicycle rentals information at an agency in Washington DC., for the years 2011-2012. The dataset also contains weather data for the same days as the bicycle rentals.<br><br>The dataset contains 731 daily records as well as 17379 hourly records for bicycle rentals at one agency in the city.<br><br>There are 17 features in the hourly dataset, and they are listed here:<br><b>instant</b> - An index for the record<br><b>dteday</b> - The date<br><b>season</b> - The prevailing climatic season<br><b>yr</b> - Year<br><b>mnth</b> - Month<br><b>hr</b> - The hour of observation<br><b>holiday</b> - Whether or not the particular day was a holiday(weekends are counted as holidays too)<br><b>weekday</b> - The day of the week<br><b>workingday</b> - Whether or not the particular day was a working day(inverse of the holiday value)<br><b>weathersit</b> - What kind of weather prevailed on the particular day<br><b>temp</b> - The hourly temperature<br><b>atemp</b> - The average temperage for the day<br><b>hum</b> - The humidity on the particular day<br><b>windspeed</b> - The windspeed on the particular day<br><b>casual</b> - The number of rentals by non-registered customers<br><b>registered</b> - The number of rentals by registered customers<br><b>cnt</b> - Total number of rentals</span>

<span style="color:red">The variables <i>weathersit</i>, <i>season</i>, <i>holiday</i> and <i>workingday</i> are categorical variables.<br><b>weathersit</b> has four states that I have simplified into: <i>clear</i>, <i>misty</i>, <i>light_precipi</i>(light rain/snow) and <i>heavy_precipi</i>(heavy rain/sleet/snow).<br><b>season</b> has four states: <i>spring</i>, <i>summer</i>, <i>fall</i> and <i>winter</i>.<br><b>holiday</b> and <b>workingday</b> are categorical variables that have binary values and orthogonal with each other in each observation.<br><br>Other observations:<br>1) Most bicycle rentals are made by registered customers than casual customers.<br>2) More bicycle rentals are made by people that are commuting to their jobs/schools than recreational customers.<br>3)Bicycle rentals are higher in dry weather than wet weather.</span>

<span style="color:black"><b><i>What is/are the main feature(s) of interest in your dataset?</i></b></span><br>
<span style="color:red">The main feature of interest in the dataset is the bicycle rental count. I am trying to build a prediction model for bicycle rentals. From our univariate analyses, we have seen that the general weather situation and season have pretty good predictive power.</span>


<span style="color:black"><b><i>What other features in the dataset do you think will help support your investigation into your feature(s) of interest?</i></b></span><br>
<span style="color:red">We need to analyse some more to build a prediction model with higher confidence.<br>I suspect windspeed, humidity and temperature will play a role in predicting bicycle rentals, but we shall see.</span>

<span style="color:black"><b><i>Did you create any new variables from existing variables in the dataset?</i></b></span><br>
<span style="color:red">No, I did not. Had the total count of bicycle rentals been missing, I could have used "casual" and "registered" to derive the total count.</span>

<span style="color:black"><b><i>Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?</span></i></b><br>
<span style="color:red">I log-transformed the total bicycle rentals for each day, to get a sense for the distribution.<br>1) When I looked at the distribution of total bicycle rentals as a whole, I found that the lower numbers occurred far more frequently. There was a tiny gap in the frequencies around the mark, which remains unexplained.<br>2) When I plotted the busiest hours of the day overall, I saw two peaks, which corresponded to the commute hours.<br>3) When I looked at the mean trends on a daily basis, I see that registered users rented mainly during the work week and lesser on weekends. For casual users, this trend was reversed - casual users rented less during the work week than on weekends.<br><br>I did not have to do anything to clean the dataset.</span>

# Bivariate plots section

<span style="color:black">From the univariate analyses performed so far, I have seen that some features tend to correlate with each other. For example, average temperatures and precipitation seem to correlate with the season in a broader sense; these variables also seem to correlate with the weather situation in a much more specific sense.</span>
```{r, echo=FALSE}
set.seed(2050)
df_sample <- df[sample(1:length(df$cnt), 5000),]
colnames(df_sample)[7] <- "hday"
colnames(df_sample)[8] <- "wday"
colnames(df_sample)[10] <- "wthr"
colnames(df_sample)[14] <- "wind"
options(warn=-1)
ggpairs(df_sample, columns=c(3,4,5,6,7,8,10,12,13,14,15), params = c(shape = I('.'), outliers.shape = I('.')), axisLabels = "none")
```
<span style="color:black"><br><br><br>A quick look at the correlation data shows that bicycle rentals are mostly influenced by the feeling temperature and hour of the day.<br><br>I will now explore some scatter plots involving total bicyle rentals together with normalized daily feeling temperatures, humidity and windspeed.</span>

```{r, echo=FALSE}
ggplot(df, aes(x=atemp, y=cnt)) + geom_point(alpha = 0.1) + geom_jitter() + geom_smooth(method = "loess") + ggtitle("Scatterplot of total bicycle rentals Vs normalized feeling temperature")
```
<span style="color:black"><br><br>From the preceding plot, we can see that bicycle rentals gradually climbs up as the temperature increases, and is lowest at the extremeties, ie., when the feeling temperatures are at their lowest and highest.</span>

```{r, echo=FALSE}
ggplot(df, aes(x=hum, y=cnt)) + geom_point(alpha = 0.1) + geom_jitter() + geom_smooth(method = "loess") + ggtitle("Scatterplot of total bicycle rentals Vs humidity")
```
<span style="color:black"><br><br>From the preceding plots, we see that most bicycle rentals occur when the normalized humidity is between 0.30-0.80.Taking a closer look at the rental number distributions between normalized humidities of 0.75-1.0, we see that there are certain "pockets" of distributions. At this point, it is not entirely clear what might be causing these distributions. It may be because certain values of normalized humidity never occur.</span>

```{r, echo=FALSE}
ggplot(df, aes(x=windspeed, y=cnt)) + geom_point(alpha = 0.1) + geom_jitter() + geom_smooth(method = "loess") + ggtitle("Scatterplot of total bicycle rentals Vs windspeed")
```

<span style="color:black"><br><br>From the preceding plots, we see that bicycle rentals are pretty steady at lower windspeeds, dropping off at higher windspeeds.</span>

```{r, echo=FALSE}
#Change values of certain columns from mere numbers to meaningful words
df$season[df$season == 1] <- "spring"
df$season[df$season == 2] <- "summer"
df$season[df$season == 3] <- "fall"
df$season[df$season == 4] <- "winter"

df$weathersit[df$weathersit == 1] <- "clear"
df$weathersit[df$weathersit == 2] <- "misty"
df$weathersit[df$weathersit == 3] <- "light_precipitation"
df$weathersit[df$weathersit == 4] <- "heavy_precipitation"
ggplot(aes(x=atemp, y=cnt, fill=weathersit), data=df) + geom_bar(stat="identity", binwidth=0.01) + ggtitle("Total bicycle rentals Vs normalized feeling temperatures")
```

<span style="color:black"><br><br>From the preceding plot, we see that bicycle rentals mostly happen in "moderate" feeling temperature. We see that clear weather rentals occur across the spectrum of feeling temperature. We also see that bicycle rentals in misty and light precipitation weather situations happen in a much narrower band of temperature. A peak stands out at a normalized feeling temperature of about 0.63, and we see that the rentals spike in clear, light precipitation and misty weather. Of course, very few rentals occur during periods of heavy precipitation.</span>

```{r, echo=FALSE}
ggplot(aes(x=hum, y=cnt, fill=weathersit), data=df) + geom_bar(stat="identity") + ggtitle("Total bicycle rentals Vs humidity")
```

<span style="color:black"><br><br>From the preceding plot, we see that most of the humidity range is dominated by clear weather and this is where a bulk of bicycle rentals take place. We do see a few peaks towards the higher end of the humidity spectrum, and this quartile seems to be dominated by misty weather for the most part and precipitative weather to a lesser extent.</span>

```{r, echo=FALSE}
ggplot(aes(x=windspeed, y=cnt, fill=weathersit), data=df) + geom_bar(stat="identity") + ggtitle("Total bicycle rentals Vs windspeed")
```

<span style="color:black"><br><br>From the preceding plot, we see that most of the bicycle rentals happen in the first two quartiles of normalized windspeed. Rentals are highest when the normalized windspeed is 0, in fact. Beyond a normalized windspeed of 0.30, however, bicycle rentals fall away steadily and are almost non-exixtent in very windy weather. Most of the visible spectrum is dominated by clear weather.<br><br>In fact, most of the bicycle rentals happen in clear weather, as shown by the summary below.</span>

```{r, echo=FALSE}
do.call(rbind , by(df$cnt, df$weathersit, summary))
```

<span style="color:black"><br><br>Let's also look at a summary of bicycle rentals by season.</span>

```{r, echo=FALSE}
do.call(rbind, by(df$cnt, df$season, summary))
```

<span style="color:black"><br>From the above summary, it looks like bicycle rentals are high in the fall and summer and taper off in the winter and spring.<br><br>Let's now look at the number of casual bicycle rentals per season.</span>

```{r, echo=FALSE}
df_master$season[df_master$season == 1] <- "spring"
df_master$season[df_master$season == 2] <- "summer"
df_master$season[df_master$season == 3] <- "fall"
df_master$season[df_master$season == 4] <- "winter"
df_master$weathersit[df_master$weathersit == 1] <- "cl"
df_master$weathersit[df_master$weathersit == 2] <- "mi"
df_master$weathersit[df_master$weathersit == 3] <- "light"
df_master$weathersit[df_master$weathersit == 4] <- "heavy"
ggplot(aes(x=season, y=casual), data=df_master) + stat_summary(fun.y=mean, geom="bar", fill="#006633") + ggtitle("Casual bicycle rentals by prevailing season")
ggplot(aes(x=season, y=registered), data=df_master) + stat_summary(fun.y=mean, geom="bar", fill="#006633") + ggtitle("Registered bicycle rentals by prevailing season")
```

<span style="color:black"><br>Though the number of registered riders using the service is consistently higher than the number of casual renters, we see that the general trends remain the same. Registered users seem to be a little more loyal than casual users, in that the winter ridership is about the same as the summer ridership - whereas it is much lower in the case of casual users.<br><br>We have already seen that average feeling temperatures and hour of the day are the main influencers of bicycle rentals. Let's look at how different weather conditions vary with the categorical variables.<br><br>We already have a feel for how feeling temperatures are connected with weather conditions. We will see how average feeling temperatures vary by season.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=atemp), data=df_master) + geom_boxplot(fill="maroon") + ggtitle("Normalized feeling temperatures by season")
do.call(rbind, by(df$atemp, df$season, summary))
```

<span style="color:black"><br>The feeling temperatures are highest in fall and summer, followed by spring and winter. This follows the general trend of bicycle rentals by season. Let's look at the trend of humidity by seasons.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=hum), data=df_master) + geom_boxplot(fill="maroon") + ggtitle("Humidity by season")
do.call(rbind, by(df$hum, df$season, summary))
```

<span style="color:black"><br>Humidity values are highest in winter, about the same in fall and summer, and lowest in spring. This is contrary to the trends we see for bicycle rentals. This reinforces our finding that humidity is not a great predictor of bicycle rentals. Let us take a look at how windspeeds influence bicycle rentals.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=windspeed), data=df_master) + geom_boxplot(fill="maroon") + ggtitle("Windspeed by season")
do.call(rbind, by(df$windspeed, df$season, summary))
```

<span style="color:black">Windspeeds are highest in spring and summer, and about the same in fall and winter. This trend is contrary to the bicycle rental trends.<br><br>From the preceding boxplots, we see that only average feeling temperature trends mimic that of bicycle rentals.</span>

```{r, echo=FALSE}
reg <- lm(formula = cnt ~ registered, data = df_master)
summary(reg)

reg <- lm(formula = cnt ~ temp, data = df_master)
summary(reg)

reg <- lm(formula = cnt ~ hr, data = df_master)
summary(reg)

reg <- lm(formula = cnt ~ season, data = df_master)
summary(reg)
```

# Bivariate analysis

<span style="color:black"><b><i>Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?</i></b></span><br>
<span style="color:red">The number of bicycle rentals correlate strongly with the year and hour of the day. The former can be attributed to the assumption that the popularity of the rental service increased in the second year. This assumption is inconclusive because we only have data from two years. I do not see any particularly strong relationships between the number of bicycle rentals and other features in the dataset like humidity, weather temperature, feeling temperature, etc., (I consider those relationships that have an R^2 value > 0.50 to be a strong relationship). <br>From the exploration that I conducted into the data, I see that about 16% of total bicycle rentals on a given day can be predicted by the prevailing temperature on that day.<br>I also see that a whopping 94% of total bicycle rentals on a given day can be predicted by the number of users that are registered with the service as on that day.<br>Given the preceding findings, I looked to see if there is a strong relationship between the number of casual bicycle renters and any other feature in the dataset. The strongest correlation I found was between casual ridership and normalized temperature - roughly 21% of casual bicycle rentals may be predicted by looking at the normalized prevailing temperatures for the day. </span>

<span style="color:black"><b><i>Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?</i></b></span><br>
<span style="color:red">I looked at the relationships between some more features in the dataset. I see that there is a moderate relationship between humidity and the prevailing weather situation of the day. I see that there is a strong relationship between the normalized temperature on a given day and the corresponding season.</span><br>

<span style="color:black"><b><i>What was the strongest relationship you found?</i></b></span><br>
<span style="color:red">The strongest relationship I found was between the total bicycle rental numbers and number of registered renters. The other strong relationship I found was between the prevailing daily normalized temperatures and the season.</span><br>

# Multivariate Plots Section

```{r, echo=FALSE}
ggplot(aes(x=weekday, y=cnt, color=season), data=df_master) + scale_x_continuous(breaks=-1:7) + stat_summary(fun.y=median, geom="line") + ggtitle("Total daily bicycle rentals by prevailing season")
```

<span style="color:black">We see that daily median ridership for each day of the week is lowest in spring and highest in the fall seasons. The daily medians numbers follow a closer trend early in the week in summer and winter.</span>

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=season), data=df_master) + scale_x_continuous(breaks=-1:24) + stat_summary(fun.y=median, geom="line")  + ggtitle("Total hourly bicycle rentals by prevailing season")
```

<span style="color:black">Looking at the median hourly ridership numbers, we see that the numbers are lowest in spring, while they follow a closer trend in the other seasons. The numbers are pretty much the same in the early hours of the day(owing to the extremely low rentals). For hourly numbers in summer, fall and winter, we see they closely converge at around 9:00am and between 4:00pm and 5:00pm - showing that these are primarily planned rentals.</span>

```{r, echo=FALSE}
ggplot(aes(x=weekday, y=cnt, color=weathersit), data=df) + scale_x_continuous(breaks=-1:7) + stat_summary(fun.y=median, geom="line")  + ggtitle("Total daily bicycle rentals by prevailing weather situation")
```

<span style="color:black">The preceding plot shows the median daily bicycle rentals depending on the prevailing weather situation of the day. Heavy precipitation seems to not deter bicycle renter on Mondays and Tuesdays, while on other days, heavy precipitation ensures low ridership. As predictable, highest bicyle ridership happens on clear days, followed by mild weather days and then on days that see light precipitation.</span>

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=weathersit), data=df) + scale_x_continuous(breaks=-1:24) + stat_summary(fun.y=median, geom="line") + ggtitle("Total hourly bicycle rentals by prevailing weather situation")
```

<span style="color:black">Lastly, we look at median hourly bicycle rentals by weather situation. As predictable, heavy precipitation ensures lowest bicycle ridership except, curiously, in the earliest hours of the day when it is higher than in any other weather situation. Up until 9:00am, we see that ridership in clear weather and mild weather are almost the same. After 9:00am, rental numbers in clear weather are higher than rental numbers in mild weather. Bicycle rentals on days with light precipitation are higher than in heavy precipitation, but lower than in clear and mild weather, at all hours of the day.</span><br>

<span style="color:black">These plots show how the hourly and daily bicycle rentals depending on the weather situation and the seasons. I wonder how these trends would vary with temperature, humidity and windspeed factors too.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=cnt), data=df) + geom_boxplot(fill="maroon") + ggtitle("Total bicycle rentals by prevailing season")
do.call(rbind, by(df$cnt, df$season, summary))
```

```{r, echo=FALSE}
ggplot(aes(x=weathersit, y=cnt), data=df) + geom_boxplot(fill="maroon") + ggtitle("Total bicycle rentals by prevailing weather situation")
do.call(rbind, by(df$cnt, df$weathersit, summary))
```

<span style="color:black">The preceding boxplots reinforce our findings that bicycle rentals are lowest in spring, followed by winter and summer and highest in the fall. Looking at the stats for each quartile, we see that the trends are close for the other three seasons except spring. We also see that periods of precipitation show lower bicycle rentals in the initial quartiles than for higher quartiles. <b><i>The main anomaly we see is that for the first quartile, bicycle rentals are higher in periods of heavy precipitation than in periods of light precipitation.</i></b>Towards the higher quartiles, though, we see that bicycle rentals in non-precipitative weather dominates.<br><br>Since bicycle rentals are dominated by registered users, and because we saw from prior analysis that bicycle rental trends closely match office/school-going patters, I thought that it might be a good idea to explore the trends among casual users. I added a new column to the dataframe to represent the ratio of casual user rentals to the total renters.</span>

```{r, echo=FALSE}
df_master$casprop <- (df_master$casual + 1)/(df_master$cnt + 1)
df_master$season <- factor(df_master$season)
df_master$weathersit <- factor(df_master$weathersit)
ggplot(data=df_master, aes(casprop, fill=season)) + geom_bar() + ggtitle("Proportion of casual users to total rentals by prevailing season")
```

<span style="color:black">We see that the bulk of ratio frequencies is below 1:2 casual:total rentals. Of these, we see that casual ridership is highest in winter, followed by summer, spring and fall. I think this may be because winter ridership may be primarily ad-hoc, while in better weather, people make plans and register for bicycle rentals.</span>

```{r, echo=FALSE}
ggplot(data=df_master, aes(casprop, fill=weathersit)) + geom_bar() + ggtitle("Proportion of casual users to total rentals by prevailing weather situation")
```

<span style="color:black">The highest occurences of casual ridership is seen in mild and light precipitative weather, followed by clear weather. There is very little casual ridership in heavy precipitation. This may be because casual riders are those who rent bicycles ad-hoc in slightly inclement weather, while they make plans and register for the service when clear weather is expected.</span>

```{r, echo=FALSE}
ggplot(data=df_master, aes(casprop, fill=as.factor(weekday))) + geom_density(alpha=0.5) + ggtitle("Proportion of casual users to total rentals by day of the week")
```

<span style="color:black">On plotting the frequency of casual ridership ratio to the total by day of the week, we see that casual ridership is higher earlier in the week than later. Interestingly, we see that casual ridership is higher on Sundays than on Saturday.</span>
```{r, echo=FALSE}
ggplot(data=df_master, aes(casprop, fill=as.factor(holiday))) + geom_density(alpha=0.5) + ggtitle("Proportion of casual users to total rentals by holiday/working day")
```

<span style="color:black">Finally, on plotting the frequencies of casual ridership ratios to the total by the fact that the days are holidays/working days, we see that casual ridership is higher on working days than on holidays. We see that on holidays, we see higher ratios of casual riders to total rentals. This may be explained by the fact that people like to explore the city on holidays.<br></span>

<span style="color:black">All of these plots reinforce the findings that we reached using earlier histograms and plots. Now, in order to build a linaear model to predict bicycle ridership, I'm going to decide on a set of variables that correlate with ridership.</span>


```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=season), data=df_master) + scale_y_log10() + scale_x_continuous(breaks=-1:24) + scale_color_brewer(type='qual') + geom_point() + ggtitle("Scatter plot of hourly bicycle rentals by the prevailing season")
```

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=weathersit), data=df_master) + scale_y_log10() + scale_x_continuous(breaks=-1:24) + scale_color_brewer(type='qual') + geom_point() + ggtitle("Scatter plot of hourly bicycle rentals by prevailing weather situation")
```

```{r, echo=FALSE}
ggplot(aes(x=atemp, y=casual), data=df_master) + geom_bar(fill="#006633", stat="identity") + ggtitle("Casual rentals Vs prevailing temperature")
```

```{r, echo=FALSE}
ggplot(aes(x=atemp, y=registered), data=df_master) + geom_bar(fill="#006633", stat="identity") + ggtitle("Registered bicycle rentals Vs prevailing temperature")
```

```{r, echo=FALSE}
ggplot(aes(x=hum, y=casual), data=df_master) + geom_bar(fill="#006633", stat="identity") + ggtitle("Casual bicycle rentals Vs humidity")
```

```{r, echo=FALSE}
ggplot(aes(x=hum, y=registered), data=df_master) + geom_bar(fill="#006633", stat="identity") + ggtitle("Registered bicycle rentals Vs humidity")
```

```{r, echo=FALSE}
ggplot(aes(registered), data=df_master) + geom_bar(fill="#006633", binwidth=5) + ggtitle("Frequency plot of individual registered rentals")
```

```{r, echo=FALSE}
ggplot(aes(x=windspeed, y=cnt), data=df_master) + geom_bar(fill="#006633", stat="identity") + ggtitle("Total bicycle rentals Vs windspeed")
```

<span style="color:black">From the plots above and the r^2 values of the linear models above, I can deduce that the features that may help us predict total bicycle rentals on any given day are "registered", "temp", "hr", "season", "weathersit", in the order of confidence. Let's try to build a linear prediction model using these features and see if we can predict bicycle rental numbers with a high level of confidence.</span>

```{r, echo=FALSE}
m1 <- lm(I(cnt) ~ I(registered), data=df_master)
m2 <- update(m1, ~ . + temp)
m3 <- update(m2, ~ . + hr)
m4 <- update(m3, ~ . + season)
m5 <- update(m4, ~ . + weathersit)
m6 <- update(m5, ~ . + windspeed)
mtable(m1, m2, m3, m4, m5, m6)
```

<span style="color:black">To the variables I mention in the section above, I added the "windspeed" variable to see if factoring it in would give my prediction model some more accuracy. We see that the prediction model I just built can help predict bicycle rentals on a given day with about 95.5% confidence.</span>

# Multivariate Analysis

<span style="color:black"><b><i>Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?</i></b></span><br>
<span style="color:red">Many of the variables interact and strengthen each others' influence over bicycle rental numbers. For example, temperature, humidity and windspeed are tightly related to the weather situation. Each season has a weather situation that dominates a majority of the days in that season.</span>

<span style="color:black"><b><i>Were there any interesting or surprising interactions between features?</i></b></span><br>
<span style="color:red">I did not see any surprising relationships between features in the dataset.</span>

<span style="color:black"><b><i>Did you create any models with your dataset? Discuss the strengths and limitations of your model.</i></b></span><br>
<span style="color:red">Yes, I created a prediction model with the dataset. The main strength of my prediction model is that it is able to predict bicycle rental numbers with about 95.5% confidence. Some limitations of the model: <br>1) The dataset spans only about two years. As we collect more and more data, we may need to tweak the model.<br>2) The model does not take into account any special events happening in the city on a given day(this is actually a limitation of the dataset!). If we know whether there were any significant events in the city(rally, game, etc.) on a given day, we could have used that as an input into our model.</span>

# Final Plots and Summary

## Plot One

```{r, echo=FALSE}
ggplot(aes(x=yr, y=cnt), data=df_master) + geom_bar(fill="#006633", stat="identity") + ggtitle("Total users year-over-year")
```

```{r, echo=FALSE}
ggplot(aes(x=yr, y=casual), data=df_master) + geom_bar(fill="#006633", stat="identity") + ggtitle("Casual users year-over-year")
```

```{r, echo=FALSE}
ggplot(aes(x=yr, y=registered), data=df_master) + geom_bar(fill="#006633", stat="identity") + ggtitle("Registered users year-over-year")
```

## Description One

<span style="color:black">From the preceding plots, we see that there is definitely an uptick in the number of bicycle rentals year-over-year. We also see that the gap between casual and registered users increases year-over-year. Therefore, we can safely say that the popularity of the bicycle rental service has clearly from the initial year to the next.</span>

## Plot Two

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=season), data=df_master) + scale_x_continuous(breaks=-1:24) + stat_summary(fun.y=median, geom="line")  + ggtitle("Total hourly bicycle rentals by prevailing season")
```

## Description Two

<span style="color:black">In the previous plot, we see the hourly trends for bicycle rentals across all the observations in the dataset, colored by seasons. In all seasons, we see that one of the hours peaks occurs at 8:00 AM and the second one at 5:00 PM. These peaks suggest that bicycle rentals are utilized by users who use the bikes to get to work/school(since these activities tend to be between fixed times everyday). We see that the cicycle rental numbers fall on either side of the peak, marking their lowest points during the night hours. The general trend holds in all seasons, but the numbers are highest in the fall, followed by summer, winter and spring. I think this is mainly because of temperature, precipitation and other factors associated with the seasons.</span>

## Plot Three

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=weathersit), data=df_master) + scale_x_continuous(breaks=-1:24) + stat_summary(fun.y=median, geom="line")  + ggtitle("Total hourly bicycle rentals by prevailing weather situation")
```

## Description Three

<span style="color:black">In the previous plot, we see hourly trends for bicycle rentals across all observations in the dataset, colored by the prevailing weather situation. The trends we see in the previous plot are, to some extent, seen in this plot too. Let's consider the three weather situations - clear, mild and light precipitation first. We see the conventional "commuter peaks" at 8:00 AM and 5:00PM, with ridership falling off on either sides of the peaks. Interestingly, renter behavior is very different during periods of heavy precipitation. Total rentals during heavy precipitation stays close to none in the early hours of the day, peaking slightly at around 6:00 PM.</span>

# Reflection

<span style="color:black">The bicycle rentals dataset consists of 17379 hourly observations of bicycle rentals at a rental agency in Washington DC. I chose this dataset because I wanted to gain some experience in choosing a tidy dataset to explore. I see that it is mostly weather-related features that closely influence the bicycle rental numbers on a given day. Factors like temperature, precipitation and windspeed tie closely into the prevailing weather situations and seasons. Therefore, to some extent, these factors influence and strengthen each other. Another feature that influences bicycle rentals is the hour of the day. I think this may be driven by office/school-goers who use bicycles to commute - and I see peaks occuring at the conventional working hours. Also, as expected, we see lesser bicycle rentals during the cold season and during periods of heavy precipitation.<br>I was able to put together a linear prediction model for bicycle rentals. The model is able to predict bicycle rentals with a 95% confidence level.<br>In conclusion, I would like to correlate this data with events that happened in Washington DC. on each of these days. That would help me see if casual bicycle rentals were driven by people wanting to avoid event-driven crowds on the popular transportation systems. also, I would love to explore the data for a few more years than contained in this dataset.</span>